Canonical rings of $\bq$-divisors on Minimal Rational Surfaces

### 2015-08-04 surfaces-write-up/surface-write-up.tex
added space
read through and made some changes

\section{Introduction}
\label{sec:intro}
For any Weil $\bq$-divisor $D$ on a rational surface $X$, the graded
\textbf{canonical ring} is $R(X, D) := \bigoplus_{d \geq 0} H^0(X, \lfloor dD \rfloor)$.
We study the generators and relations of canonical
rings for projective spaces $\bp^m$ and Hirzebruch surfaces $\hirz_m$.
When $D$ is a general $\bq$-divisor on any $\bp^m$ or on any
$\hirz_m$, we give bounds on the generators and relations of
$R(X, D)$. In particular, we give a complete description of the
canonical ring when $X = \bp^m$ and $D$ is any effective
$\bq$-divisor.

In part, this work is motivated by a recent paper by
O'Dorney \cite{dorney:canonical}, which gives similar descriptions
of canonical rings for general $\bq$-divisors on $\bp^1$. This
study also has connections to the papers by Voight--Zureick-Brown 
\cite{vzb:stacky} and Landesman--Ruhm--Zhang \cite{lrz:spin-cring}
on log spin canonical rings on stacky curves which provide bounds
on rings of modular forms of Fuchsian groups through the study
of particular $\bq$-divisors on curves.

The work in this paper may help calculate a presentation of certain rings of
Hilbert 
modular forms and Siegel modular forms. Recall that Hilbert and Siegel 
modular forms are both generalizations of modular forms to higher dimensions,
as described in \cite{geer:siegel-modular} and \cite{bruinier:hilbert-modular}.
By a result by Voight and Zureick-Brown \cite[Proposition 6.1.5]{vzb:stacky}, 
one can translate the analytic category of 1-dimensional complex orbifolds to
the algebraic category of stacky curves. 
We expect that one can generalize this result to yield an equivalence of
categories between an appropriate subcategory of $m$-dimensional orbifolds and
a subcategory of $m$-dimensional algebraic spaces.
In particular, this would allow us to translate rings of Hilbert modular forms
and Siegel modular forms whose coarse space is a projective space or a
Hirzebruch surface to canonical rings on the corresponding algebraic surfaces.
Thereby, our techniques would give a bound on generators and relations for
Hilbert and Siegel modular forms with certain underlying spaces.

@@

In section 2, we discuss background knowledge and present some useful lemmas.  In section 3 we focus on projective spaces: $\mathbb{P}^n$ and prove the following two results:
\begin{restatable}{thm}{restateeffective}
\label{thm:proj-effective-intro}
Let $D = \sum_{i = 1}^{n} \alpha_i C_i \in \bz \otimes_\bz \di \bp^m$, with $\alpha_i =
\frac{p_i}{q_i} \in \bq$ and each $C_i$ a hypersurface of
$\bp^m$, be an effective divisor on $\bp^m$.

@@

\noindent
In section 4, we shift our attention to Hirzebruch Surfaces and prove the following theorem:

@@

\noindent
Together, Theorems \ref{thm:proj-generators-relations} and \ref{thm:hirz-generators-relations} provide bounds on generators and relations for the canonical ring of $\mathbb{Q}$-divisors of all minimal rational surfaces.  Section 5 concludes the paper and discusses further questions.

@@

Now we give a result describing the generators of a cone via its
extremal rays.
\todo{write more transition}

@@

\section{Canonical Rings on Projective Space}
\label{sec:proj}
In this section, we bound the degrees of generators and relations for divisors
on $\bp^m$, for all $m \geq 1$. The $\bp^1$ case was covered in
\cite{dorney:canonical}.

@@

For the remainder of this section, we shall fix $m \geq 1$ and choose an
isomorphism $\bp^m \cong \proj V$ so that $x_0,\ldots, x_m$ form a basis for
$V$.  We will think of $x_i$ as a rational section of $H^0(\bp^m,
\sco_{\bp^m})$ of degree 1.

@@

\begin{prop}
\label{prop:pm-span-and-basis}
The functions $u^d \cdot \prod_{i=1}^n f_i^{c_i}$ satisfying
\begin{align}
\label{align:pm-span}
\sum_{i=1}^{n} c_i \cdot \bida_i = 0 && \text{ and } &&c_i \geq \lfloor \alpha_i d\rfloor	
\end{align}

\noindent
form a spanning set for $H^0(\bp^m, dD)$ over $k$. Furthermore, functions 
of the form $u^d \cdot \prod_{i=1}^n f_i^{c_i}$ satisfying
\begin{align}
\label{align:pm-basis}
c_i = -\lfloor d\alpha_i \rfloor && \text{ for } && i > m
\end{align}

@@

In this subsection, we restrict attention to the case of effective
fractional divisors $D \in \bq \otimes \di \bp^m$. We completely
characterize the canonical rings when $D$ consists of one
hypersurface and give strong bounds on generator and relation
degrees for general effective divisors $D$.

\begin{convention}
Let $[x_0, \ldots, x_m]$ be coordinates for $\bp^m$. Let
$\vec{v} = (v_0, \ldots, v_m) \in \bq^{m + 1}$. Then write
\[
	x^{\vec{v}} := \prod_{i = 0}^{m} x_i^{v_i}
\]
\end{convention}

@@

\noindent
has a minimal presentation over $R_{D'}$ consisting of the $\sum_{i = 0}^{r}
{{m + c_i - 1} \choose {c_i}}$ generators $F_i^{\vec{v}} := \frac{u^{d_i}
x^{\vec{v}}}{x_k^{c_i}}$ where $1 \leq i \leq r$, $\vec{v} \in \bz_{\geq 0}^{m + 1}$
with $\deg \vec v = c_i$. Furthermore, it has
relations generated by the following two forms:
\begin{enumerate}
	\item For each $(i, j)$ with $j \geq i + 2$ and each $\vec{v} \in \mss_i,
\vec{w} \in \mss_j$, there is a relation of the form either
\begin{flalign*}
	&G_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- \prod_{\vec{y} \in \mss_{h_{i, j}}} (F_{h_{i, j}}^{\vec{y}})
	^{g_{\vec{y}}} &(i < h_{i, j} < j) \\
	\text{or} \\
	&G_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- \prod_{\vec{y} \in \mss_{h_{i, j}}} (F_{h_{i, j}}^{\vec{y}})^{g_{\vec{y}}}
	\cdot \prod_{ \vec{z} \in
	\mss_{h_{i, j} + 1}} (F_{k_{i, j} + 1}^{\vec{z}})^{g'_{\vec{z}}}
	&(i < h_{i, j} < h_{i, j} + 1 < j).
\end{flalign*}
	\item For each $(i, j)$ with
$j = i$ or $j = i + 1$ and each $\vec{v} \in \mss_i, \vec{w} \in
\mss_j$ with $\vec{v} \not\prec \vec{w}$ (see Definition
~\ref{defn:vec-order}) there is a relation of the form
\begin{align*}
	&L_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- F_i^{\vec{y}} F_j^{\vec{z}} \\
\end{align*}

@@

\noindent
Therefore, $\ker \psi$ is generated in degreesat most
\[
	\tau = \rho
	+ \max \left(\max_{i\in \Te} \bigl(\ell_i \gcd(a_i, b_i) \bigr),
	\; \max_{\substack{i \in \Tp \\ j \in \Tm}} \bigl(\ell_{i, j}
	(\bida_i \bidb_j - \bida_j
	\bidb_i) \bigr) \right).
\]
\end{proof}

@@

Answering Question ~\ref{qn:general-blowup} would in fact provide an answer for
all rational surfaces, via the minimal model program for surfaces
\todo{Reference}.  One might then hope to further extend these results to all minimal surfaces:
\begin{question}
\label{qn:general-minimal-surface}
Can we describe generators and relations of $R_D$ for $D$ a divisor on an
arbitrary minimal surface $X$?
\end{question}

>>>>

\section{Introduction}
\label{sec:intro}
For any Weil $\bq$-divisor $D$ on a rational surface $X$, the graded
\textbf{canonical ring} is $R(X, D) := \bigoplus_{d \geq 0} H^0(X, \lfloor dD \rfloor)$.
We study the generators and relations of canonical
rings for projective spaces $\bp^m$ and Hirzebruch surfaces $\hirz_m$.
When $D$ is a general $\bq$-divisor on any $\bp^m$ or on any Hirzebruch Surface
$\hirz_m$, we give bounds on the generators and relations of
$R(X, D)$. In particular, we give a complete description of the
canonical ring when $X = \bp^m$ and $D$ is any effective
$\bq$-divisor.

In part, this work is motivated by a recent paper by
O'Dorney \cite{dorney:canonical}, which gives similar descriptions
of canonical rings for general $\bq$-divisors on $\bp^1$. This
study also has connections to the papers by Voight--Zureick-Brown 
\cite{vzb:stacky} and Landesman--Ruhm--Zhang \cite{lrz:spin-cring}
on canonical rings of stacky curves which provide bounds
on rings of modular forms of Fuchsian groups through the study
of particular $\bq$-divisors on curves.  \todo{Peter: I think this is too much specificity about our result.  I think everything after ``stacky curves'' should be eliminated, as it seems a little self-promoting.}

The work in this paper may help calculate a presentation of certain rings of
Hilbert 
modular forms and Siegel modular forms. Recall that Hilbert and Siegel 
modular forms are two generalizations of modular forms to higher dimensions,
as described in \cite{geer:siegel-modular} and \cite{bruinier:hilbert-modular}.
By Voight and Zureick-Brown \cite[Proposition 6.1.5]{vzb:stacky}, 
one can translate the analytic category of 1-dimensional complex orbifolds to
the algebraic category of stacky curves. 
We expect that one can generalize this result to yield an equivalence of
categories between an appropriate subcategory of $m$-dimensional orbifolds and
a subcategory of $m$-dimensional algebraic spaces.
In particular, this would allow us to translate rings of Hilbert modular forms
and Siegel modular forms whose coarse space is a projective space or a
Hirzebruch surface to canonical rings on the corresponding algebraic surfaces.
Thereby, our techniques would give a bound on generators and relations for
Hilbert and Siegel modular forms with certain underlying spaces.
\todo{Peter:  Hmm... this is very vague.  It almost feels like a further question type statement, then a statement of why what we're doing matters.  I guess the real issue, as we have sort of discussed, is that there are not any clear motivating reasons why one would actually use what we have done... Oh well}

@@

In Section \ref{sec:lemmas}, we discuss background knowledge and present some useful lemmas.  In Section \ref{sec:proj} we focus on projective spaces: $\mathbb{P}^n$ and prove the following two results:
\begin{restatable}{thm}{restateeffective}
\label{thm:proj-effective-intro}
Let $D = \sum_{i = 1}^{n} \alpha_i C_i \in \bz \otimes_\bz \di \bp^m$, with $\alpha_i =
\frac{p_i}{q_i} \in \bq$ and each $C_i$ a hypersurface of
$\bp^m$, be an effective divisor on $\bp^m$.

@@

\noindent
In section \ref{sec:hirz}, we shift our attention to Hirzebruch Surfaces and prove the following theorem:

@@

\noindent
Together, Theorems \ref{thm:proj-generators-relations} and \ref{thm:hirz-generators-relations} provide bounds on generators and relations for the canonical ring of $\mathbb{Q}$-divisors of all minimal rational surfaces.  Section \ref{sec:conc} concludes the paper and discusses further questions.

@@

We now present some preliminary results which will be used to prove Theorems \ref{thm:proj-generators-relations} and \ref{thm:hirz-generators-relations} in Sections \ref{sec:proj} and \ref{sec:hirz} respectively.
%Now we give a result describing the generators of a cone via its
%extremal rays.
%\todo{write more transition}

@@

\section{Canonical Rings on Projective Space}
\label{sec:proj}
In this section, we bound the degrees of generators and relations for divisors
on $\bp^m$, for all $m \geq 1$. The $\bp^1$ case restricts to the results of
\cite{dorney:canonical}.

@@

For the remainder of this section, we shall fix $m \geq 1$ and choose an
isomorphism $\bp^m \cong \proj V$ for some vector space $V$ with basis $x_0,\ldots, x_m$.  We will think of $x_i$ as a rational section of $H^0(\bp^m,
\sco_{\bp^m})$ of degree 1.

@@

\begin{prop}
\label{prop:pm-span-and-basis}
The functions $u^d \cdot \prod_{i=1}^n f_i^{c_i}$ satisfying \todo{Peter: Why do we include this $u$?  IN the later sections, I don't use this notation and we should standardize.}
\begin{align}
\label{align:pm-span}
\sum_{i=1}^{n} c_i \cdot \bida_i = 0 && \text{ and } &&c_i \geq \lfloor \alpha_i d\rfloor	
\end{align}

\noindent
form a spanning set for $H^0(\bp^m, dD)$ over $k$. Furthermore, functions 
of the form $u^d \cdot \prod_{i=1}^n f_i^{c_i}$ satisfying
\begin{align}
\label{align:pm-basis}
c_i = -\lfloor d\alpha_i \rfloor && \text{ for } && i > m
\end{align}\todo{Peter: Should be for $i> m+1$ right?  Looking below I see an $f_0$, where as above $f_i$ ranges 1 to $n$.  We should choose whether or not 0 is included in the indexing and fix the section to match that convention.}

@@

In this subsection, we restrict attention to the case of effective
fractional divisors $D \in \bq \otimes \di \bp^m$. We completely
characterize the canonical rings when $D$ consists of one
hypersurface and give strong \todo{Peter: I'm opposed to the inclusion of the word strong, as it ambiguous and subjective.  I think we should either replace it with tight, and then explicitly say later how it is tight, or just get rid of the adjective.} bounds on generator and relation
degrees for general effective divisors $D$.

\begin{convention}
Let $[x_0, \ldots, x_m]$ be coordinates for $\bp^m$. Let
$\vec{v} = (v_0, \ldots, v_m) \in \bq^{m + 1}$. \todo{Peter: do we want to allow $\vec{v}\in \mathbb{Q}^{m+1}$ rather than $\mathbb{Z}^{m+1}$?  I'm not sure if the arguments actually need the generality (and if so, it is a little worrisome since fractional powers aren't clearly defined algebraically.}  Then write
\[
	x^{\vec{v}} := \prod_{i = 0}^{m} x_i^{v_i}
\]
\end{convention}

@@

\noindent
has a minimal presentation over $R_{D'}$ consisting of the $\sum_{i = 0}^{r}
{{m + c_i - 1} \choose {c_i}}$ generators $F_i^{\vec{v}} := \frac{u^{d_i}
x^{\vec{v}}}{x_k^{c_i}}$ where $1 \leq i \leq r$, $\vec{v} \in \bz_{\geq 0}^{m + 1}$
with $\deg \vec v = c_i$. Furthermore, it has
relations generated by the following two forms:
\begin{enumerate}
	\item For each $(i, j)$ with $j \geq i + 2$ and each $\vec{v} \in \mss_i,
\vec{w} \in \mss_j$, there is a relation of the form: either
\begin{flalign*}
	&G_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- \prod_{\vec{y} \in \mss_{h_{i, j}}} (F_{h_{i, j}}^{\vec{y}})
	^{g_{\vec{y}}} &(i < h_{i, j} < j) \\
	\text{or} \\
	&G_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- \prod_{\vec{y} \in \mss_{h_{i, j}}} (F_{h_{i, j}}^{\vec{y}})^{g_{\vec{y}}}
	\cdot \prod_{ \vec{z} \in
	\mss_{h_{i, j} + 1}} (F_{k_{i, j} + 1}^{\vec{z}})^{g'_{\vec{z}}}
	&(i < h_{i, j} < h_{i, j} + 1 < j).
\end{flalign*} \todo{Peter: Can you clarify when these two cases occur: i.e. does the first only occur if $j = h_{i,j}+1$; otherwise, in cases when $i< h_{i,j} < h_{i,j} + 1< j$, do we get relations of both forms?}
	\item For each $(i, j)$ with
$j = i$ or $j = i + 1$ and each $\vec{v} \in \mss_i, \vec{w} \in
\mss_j$ with $\vec{v} \not\prec \vec{w}$ (see Definition
~\ref{defn:vec-order}) there is a relation of the form
\begin{align*}
	&L_{i, j}^{\vec{v}, \vec{w}} = F_i^{\vec{v}} F_j^{\vec{w}}
	- F_i^{\vec{y}} F_j^{\vec{z}} \\
\end{align*}

@@

\noindent
Therefore, $\ker \psi$ is generated in degrees at most
\[
	\tau = \rho
	+ \max \left(\max_{i\in \Te} \bigl(\ell_i \gcd(a_i, b_i) \bigr),
	\; \max_{\substack{i \in \Tp \\ j \in \Tm}} \bigl(\ell_{i, j}
	(\bida_i \bidb_j - \bida_j
	\bidb_i) \bigr) \right).
\]
\end{proof}

@@

Answering Question ~\ref{qn:general-blowup} would in fact provide an answer for
all rational surfaces, via the minimal model program for surfaces.  One might then hope to further extend these results to all minimal surfaces:
\begin{question}
\label{qn:general-minimal-surface}
Can we describe generators and relations of $R_D$ for $D$ a divisor on an
arbitrary minimal surface $X$?
\end{question}
